\documentclass[twocolumn,10pt]{article}

\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{listings}
\usepackage{graphicx}
\usepackage{comment}
\usepackage{amsmath}
\usepackage{hyperref}

\usepackage[a4paper, total={6.4in, 8.7in}]{geometry}

\title{Assessed Coursework Report\linebreak Advanced Operating Systems}

\author{Willian de Oliveira Barreiros Junior\\
Matric Number: 2105514\\
 guns945@gmail.com\\}

\begin{document}
\pagenumbering{gobble}
\clearpage
\thispagestyle{empty}
\maketitle

\section{Introduction}
This report regards the evaluation of the features provided by Erlang programming language by implementing a small multiprocess temperature system.

This report is divided into four parts, a brief introduction on Erlang's background and basic features, the proposed case  study (the temperature multiprocess program), the findings on Erlang's qualities and shortcomings given the experience on developing the temperature program, and finally, some final remarks concluding this work.

The source code from this coursework can also be found at
\url{https://github.com/WillianJunior/AdvancedOSExercice}

\section{The Erlang Programming Language}
First and foremost, Erlang was chosen given that it already had the message passing system built in its core and suport a programming paradigm that uses \textit{share nothing} processes with message passing. Also, it was rather helpful that Erlang is a functional programming language, which, in my experience, results in shorter and more concise code.

Erlang is, as stated before, a functional programming language focused on concurrency and reliability. For being a functional language, parallelism can be easily and safely achieved.

Erlang uses a simple mesage passing system as the only interprocess comunication tool. This can be very useful since it's easier to reason large an complex systems as a set of actors comunicating between each other using only messages instead of any kind of shared memory emulation, semaphores and locks.

Erlang proposes a concurrent oriented programming paradigm. This enables escalability and safety given that the processes are independent from each other. This also makes easy to update the system while it's still running by implementing hot code loaders.

One controversial feature is that Erlang is dynamically typed. When comparing with a statically typed language (e.g. Haskell), it's possible to see the shortcomes. It's much harder to have runtime errors with statically typed languages. This, however, is not a problem given the "let it crash" philosofy of Erlang. This approach is taken given that its impossible to predict and treat every possible error (e.g. hardware errors) and that one process should not interfere with another.

\section{Case Study Development}
As a case study, a simple multiprocess system was developed. This system had as its main goals extensibility and error tolerance.

The system itself consists on a clock process that notify a couple o sensors processes to display their readings. This was done by requesting the conversion to the converter process througth a message. The converter in turn sends a response message to the requester with the result of the conversion. Finaly, the sensors send the converted temperature to the display.

Extensibility was achieved by enabling load of new convertion functions and addition of new sensors on runtime. This was easily done with the aid of \textit{atoms}.

Since the processes have a\textit{share nothing} aspect, there isn't much that need to be done in order to achieve fault tolerance. The only extension done was to ensure that if a process fails for any reason, except \textit{normal} end of execution or a \textit{shutdown} signal, it shaw be restarted. This is performed by wrapping each process with a supervisor process and using \textit{atoms} to reference the processes.

\section{Case Study Results}
Erlang prooved itself a feasible choice as a programming language for large, extensible and safe systems, even for beginners. This is given mainly because of its learning curve and available documentation. 

The first quality is that the instalation process (on linux at least) is extreamily trivial (just as with Haskell), with the instalation package comming with everything you need.

When it comes to the learning, the closest programming language I've worked before was Haskell; consequently the functional portion of the language was trivial to learn, taking only 15 to 30 minutes to review the concepts. The first barrier encountered on the learning process was the sintax (e.g. ',' '.' ';'), which was, at least, unique for me. After that, the next step would be learning how to use the message passing system and the multiprocess system. Those were also trivial, resulting in a first draft of the case study system done within a hour, including all message passing protocol and processes.

Given that the first draft had less than 50 lines of code, other improvements were thougth to carry on with the learning process. After some reading a message passing pattern arose: encapsulating the message passing into a function of the receiving module. This would abstract the actual message protocol from the sender, resulting in a more reusable code.

The next improvement regards error tolerance. One of the most expressive characteristics of Erlang is the \textit{let it fail} philosofy. This only works because the ease to use supervisors. To test it a simple supervisor was implemented for each process, restarting it whenever it fails. The implementation was trivial, but also incomplete. A better aproach would be to extract the supervisor code into a module and make a tree of supervisors, implementing a protocol of contingencies for different fails on different levels. This was not implemented because it would increase (even more) the complexity of the case study. An important aspect to highlight is the use of \textit{atoms} to make the processes management transparent to the message passing system. E.g. let's say a process die and it's restarted. This process now have a different \textit{Pid}. This is a problem given that in order to pass a message you need the receiver \textit{Pid}. However, instead of having to implement a system to keep the \textit{Pid} references concistent \textit{atoms} were used as a kind of sefl-updating constant (in terms of other processes) \textit{Pid}. The esulting code sized arround 90 lines.

The final improvement was to enable loading of convertion functions and creation of new sensors on runtime. For this amend it was necessary knowledge on how to hold state on a process, high-order functions and even more \textit{atoms}. For the loading of functions, the converter process could also receive a \textit{new\_function} message, on which it would add  the new function to its current state. The function would come along a identifier (for lookup purposes) in a form of a \textit{atom}. The usage of \textit{atoms} proved itself very useful since it doesn't generate a large overhead on the messages (since its implemented as a number) and also makes the code easy to interpret. The dynamic creation of new sensors functionality was also not that difficult given the use of atoms as the references to the processes. The final code is arround 200 lines long, being almost half of it comments and the replicated supervisor.

\section{Conclusion}
Erlang proved itself to be simple enougth to learn the basics. The learning speed is somewhat quick given what you can acomplish comparing what you would need to learn on other more conventional languages (e.g. java or C). The size and conciseness of the code are defnetly better when comparing to the mainstream languges. The full development process, including studying and testing, took no longer that 8 hours, which is relatively a very small amount of time to actually learn a new progaming language.

Some advantages found were the facility to use the message passing built-in functions and to program conforming the paradigm of \textit{share nothing} processes. Also the fact that Erlang is functional helped a lot. One of the most interesting aspects found was \textit{atoms}, which proved to be very powerful.

Although Erlang has its advantages the fact that it is dynamicaly typed can be a source of several bugs for newbies programers. One other aspect that may seem unimportant compared to everything else are the anonymous functions. They seem quite unnatural when compared to Haskell. The syntax also is rather unusual and can take some time to get used to it.

Given the abovementioned reasons, Erlang is a viable choice for any kind of programmer that need to build a scalable and safe distributed system.

\end{document}